### 2016-06-26 store-consistency.tex
Figures added to 3.2. Edits done.

In practice, relational databases resolve the conflict by prioritizing
weak isolation (thus, concurrency and performance) over strong
consistency, so the execution traces do not necessarily satisfy {\sc
sc}. In particular, visibility constraints imposed by {\sc sc} are
violated if (and only if) they are found to be in conflict with the
constraints imposed by the isolation level. In the context of
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store needs to make $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term
\emph{recency}~\cite{bailisvldb} is often used in place of consistency
to capture the best-effort nature of the {\sc sc} guarantee on
databases.}. We generalize this approach to any $\I_s$ and $\I_c$ by
defining a \emph{maximum visiblity principle} to determine an
acceptable weakening of $\I_s$ in case of a conflict with $\I_c$.  The
principle requires the weakened consistency guarantee ($\I_s'$) of the
store to enforce all visibility relationships imposed by the actual
consistency guarantee ($\I_s$), unless enforcing such a relationship
violates $\I_c$. 
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes relational database-like stores.

@@

\begin{remark}
Note that the purpose of maximum visibility principle is to
rationalize the observed behaviour of databases in practice. In
particular, it is not intended to be a guiding principle to engineer
data stores, and need not be objectively better than similar
principles.
\end{remark}

@@

Reasoning under weakly consistent replication has received special
attention in the recent research~\cite{gotsmanpopl16}. Our operational
semantics and proof system are however more general and admit
replication as a special case. In this section, we explain how our
reasoning approach works under weakly consistent replication by
comparing our model with the standard model of replication.

@@

\GK{Here, I plan to include the full example that we dropped from
\S\ref{sec:motivation}.}

>>>>

In practice, relational databases resolve the conflict by prioritizing
weak isolation (thus, concurrency and performance) over strong
consistency, so the execution traces do not necessarily satisfy {\sc
sc}. In particular, visibility constraints imposed by {\sc sc} are
violated if (and only if) they are found to be in conflict with the
constraints imposed by the isolation level. In the context of
aforementioned example, $\eta_1$ is not made visible to $\eta_2$
because doing so would violate $\I_c$. However, if $\I_c$ is true,
then the store makes $\eta_1$ visible to $\eta_2$ to honor its
consistency commitment\footnote{The term \emph{recency
commitment}~\cite{bailisvldb} is often used in practice, perhaps to
capture the best-effort nature of {\sc sc}.}. We generalize this
approach to any $\I_s$ and $\I_c$ by defining a \emph{maximum
visiblity principle} to determine an acceptable weakening of $\I_s$ in
case of a conflict with $\I_c$.  The principle requires the weakened
consistency guarantee ($\I_s'$) of the store to enforce all visibility
relationships imposed by the actual consistency guarantee ($\I_s$),
unless enforcing such a relationship violates $\I_c$. 
% The formal definition of the principle is relegated
% to the supplementary in the interest of space, but it can be
% understood in the context of an {\sc sc} store, where it weakens the
% store invariant to the following:
Formally:
\begin{definition}
$\I_s' : \E \rightarrow \Prop$ is said to be a maximum visibility
weakening of $\I_s : \E \rightarrow \Prop$ if and only if:
\begin{itemize}
  \item $\I_s'$ is weaker than $\I_s$: 
      $\forall\E.~ \I_s(\E) \Rightarrow \I_s'(\E)$, and
  \item In every trace $\E$ that satisfies $\I_c$, and for every pair
  of effects $\eta_1$ and $\eta_2$ in $\E$, if $\I_s(\E)$ requires
  $\eta_1$ to be visible to $\eta_2$, then so does $\I_s'(\E)$ unless
  extending $\E$ with $\visZ(\eta_1,\eta_2)$ violates
  $\I_c$\footnote{\GK{ToDo: consider other well-formedness conditions
  on trace, such as acyclicity of $\visZ$ and $\soZ$. Are they needed
  (considering that the machine never violates them)? Encode the
  specifications in Z3 and make sure they are consistent.}}:
  \begin{smathpar}
  \begin{array}{l}
  \forall\E,\eta_1,\eta_2.~ \I_c(\E) \Rightarrow (\I_s(\E)
    \Rightarrow \underE{\eta_1 \visar \eta_2}) \Rightarrow \\
    \hspace*{0.5in}(\I_s'(\E) \Rightarrow \underE{\eta_1 \visar
    \eta_2} \disj \neg\I_c(\E\,\cup\,(\emptyset,\{(\eta_1,\eta_2)\})))
  \end{array}
  \end{smathpar}
\end{itemize}
\end{definition}
Applying this principle, we can weaken {\sc sc} to obtain the
following store trace invariant ($\I_s$) for an {\sc sc} store whose
isolation constraints are captured by $\I_c$:
\begin{smathpar}
\begin{array}{lcl}
\I_s(\E) & = & \forall \eta_1,\eta_2.\, \{\eta_1,\eta_2\},
    \subseteq \E.\A \conj \id(\eta_1) <
    \id(\eta_2) \\
    & & \hspace*{0.5in} \Rightarrow 
      \underE{\eta_1 \visar \eta_2} \disj \neg\I_c(\E
    \cup (\emptyset,\{(\eta_1,\eta_2)\}))\\
\end{array}
\end{smathpar}
$\I_s$ requires a trace $\E$ to satisfy the visibility constraints of
{\sc sc} except in cases where they are in conflict with $\I_c$.
Instantiating the parameter $\I$  with $\I_s \wedge \I_c$ in
Fig.~\ref{fig:txnimp} results in an operational semantics that
describes relational database-like stores.

@@

\begin{remark}
Note that the purpose of maximum visibility principle is to
rationalize the observed behaviour of databases in practice. In
particular, it is not intended to be a guiding principle to engineer
data stores.
\end{remark}

@@

Reasoning under weakly consistent replication has received special
attention in the recent work~\cite{gotsmanpopl16}. Our operational
semantics and proof system are general enough to admit replication as
a special case. In this section, we explain how our reasoning approach
works under weakly consistent replication by comparing our model with
the standard model of replication.

@@

\begin{figure}
\centering
\begin{txnimpcode}
 $\begin{decoration}
 P_1:\{ {\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj\\
        \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2}
        \hboar \C{Wd1} \wedge \C{Wd2} \wrstoar \C{C} \wedge \C{C = k-a2} \}
 \end{decoration}$
  txn$\langle$'Wd1'$\rangle${
   $\begin{decoration}
    \phi_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \conj
      {\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
        {\C{Wd2} \hboar \C{Wd1}} 
       \Rightarrow \C{C = k-a2} \}
    \end{decoration}$ 
    v1 = C
   $\begin{decoration}
    \phi_2 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v1 = k} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
       \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
        {\C{Wd2} \hboar \C{Wd1}} 
       \Rightarrow \C{C = k-a2} \wedge \C{v1 = k-a2}\}
    \end{decoration}$ 
    if (v1 $\ge$ a1) {
      v2 = C;
     $\begin{decoration}
      \phi_3 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v2 = k} \conj\\
       \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
          {\C{Wd2} \hboar \C{Wd1}} \Rightarrow \C{C = k-a2} \wedge \\
          \hspace*{1.9in}\C{v2 = k-a2}\}
      \end{decoration}$ 
      v3 = v2 - a1;
     $\begin{decoration}
      \phi_4 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k} \wedge \C{v3 = k-a1} \conj\\
         \hspace*{0.3in}{\committed(\C{Wd2})} \Rightarrow \C{Wd2} \wrstoar \C{C} \conj \\
         \hspace*{0.3in}{\committed(\C{Wd2})} \wedge
          {\C{Wd2} \hboar \C{Wd1}} 
         \Rightarrow \C{C = k-a2} \wedge \\
         \hspace*{1.9in}\C{v3 = k-a2-a1}\}
      \end{decoration}$ 
      C := v3
     $\begin{decoration}
      \phi_5 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}) 
                \conj\\
         \hspace*{0.3in}{\committed(\C{Wd2})} 
                \Rightarrow \C{C = k-a1-a2}\}
      \end{decoration}$ 
    }
  }
 $\begin{decoration}
  Q_1 : \{{\neg\committed(\C{Wd2})} \Rightarrow \C{C = k-a1}
            \conj \committed(\C{Wd1}) \conj\\
      \hspace*{0.12in}{\committed(\C{Wd2})} 
          \Rightarrow \C{C = k-a1-a2} \}
  \end{decoration}$ 
\end{txnimpcode}

\caption{`Wd1' transaction decorated with assertions}
\label{fig:wd1-decorated}
\end{figure}

The fully decorated implementation of transaction `Wd1' is shown in
Fig.~\ref{fig:wd1-decorated}. Decorated `Wd2' is similar. Assignment
statements are broken down and temporary local variables (\C{v1},
\C{v2} and \C{v3}) are introduced so as to separate shared variable
reads and writes.  Since we reason in terms of executions, all
assertions implicitly refer to the current execution (just as hoare
triples implicitly refer to the current state). Also implicit is the
invariant \C{k $\ge$ a1+a2}. Proposition $\committed(T)$ indicates
that the transaction $T$ is already committed. Precondition ($P_1$) of
`Wd1' accounts for the possibility of `Wd2' committing before `Wd1',
writing \C{k-a2} to \C{C}.  Precondition of `Wd2' is similar.  Since
neither `Wd1' nor `Wd2' are committed at the beginning, the
precondition in Fig.~\ref{fig:motiv-eg-1} is extended to include
$\neg\committed(\C{Wd1}) \wedge \neg\committed(\C{Wd2})$, from which
$P_1$ follows. Once the execution is inside `Wd1', commit of `Wd2'
($\committed(\C{Wd2})$) may mean that either `Wd2' happened before
`Wd1', or that it committed concurrently with `Wd1'. Since SI
proscribes the latter possibility, we only consider the case when
$\C{Wd2} \hboar \C{Wd1}$. A proof for $\C{Wd2} \hboar \C{Wd1}$ is
obtained subsequently, allowing us to get rid of the special case.

The first obligation is to show that the assertions that decorate
`Wd1' are indeed valid. This entails proving that, for each statement
of `Wd1', if evaluating the statement and extending an execution that
satisfies its precondition results in a well-formed execution, then
the extended execution must satisfy the postcondition. The
well-formedness assumption lets us ignore the executions that violate
the trace invariant lead the machine to a stuck state. This is useful
in case of the assignment statement that writes \C{v3} to \C{C}. Since
`Wd2' also writes to \C{C}, the trace invariant ($\I$) requires
$\C{Wd2} \hboar \C{Wd1}$ if $\committed(\C{Wd2})$ (the alternative
$\C{Wd1} \hboar \C{COMMIT}(\C{Wd2})$ allowed by \C{SI(Wd1)} is
impossible in this case because \C{Wd2} has already committed, while
\C{Wd1} is still in progress). This lets us treat the special case
$\committed(\C{Wd2})\conj \C{Wd2} \hboar \C{Wd1}$ as being equivalent
to the general case $\committed(\C{Wd2})$, which is required to prove
the postcondition of `Wd1'. Observe that \C{RC(Wd1)} does not justify
this generalization, hence replacing \C{SI(Wd1)} by \C{RC(Wd1)} is not
sufficient to prove the postcondition of `Wd1'.

The second obligation is to prove that assertions are valid despite
the interference from the the concurrent thread executing `Wd2'. The
interference is given by the following rely relation ($R_1$):
\begin{smathpar}
\begin{array}{lcl}
  R_1 & = & \{ (\E,\E') \;|\; \neg\underE{\committed(\C{Wd2})} \conj 
        \underE{\I} \conj \E'\Vdash \I \conj\\
%       \conj (\E'-\E) \subseteq \C{Wd2} \conj \\
%   & & \hspace*{0.5in} \E' \Vdash \committed(\C{Wd2}) ~\Rightarrow~ \E'
%       \Vdash \C{Wd2} \wrstoar \C{C} \conj \\ 
%   & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj \E' \Vdash
%   \committed(\C{Wd2}) \Rightarrow \C{C=k-a1-a2} \\
    & & \hspace*{0.5in} \neg\underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.8in}\Rightarrow \C{C=k-a2} \conj 
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \conj \\
    & & \hspace*{0.5in} \underE{\committed(\C{Wd1})} \conj
        \C{COMMIT(Wd2)} \in (\E'-\E) \\
    & & \hspace*{0.6in}\Rightarrow \C{C=k-a1-a2} \conj
        \E' \Vdash \C{Wd2} \wrstoar \C{C} \}\\
\end{array}
\end{smathpar}
The rely relation says the following about the concurrent thread: (1).
It may interfere only if `Wd2' is not already committed, (2). Any
interference takes well-formed executions to well-formed executions,
(3). If the interference commits `Wd2', then `Wd2' should have already
written to \C{C}, and (4). The value written is either $\C{k-a1-a2}$
or $\C{k-a1}$ depending on whether or not `Wd1' has already committed.
Interference from `Wd2' is allowed between any two operations of
`Wd1', and assertions are required to be valid despite the
interference. For instance, if the execution ($\E$) immediately after
evaluating $\C{v1 = C}$ satisfies $\phi_2$ (i.e., $\underE{\phi_2}$),
and an interference from `Wd2' extends $\E$ to $\E'$ (i.e.,
$R_1(\E,\E')$), then the execution $\E'$ just before evaluating the
condition must satisfy $\phi_2$ (i.e., $\E' \Vdash \phi_2$). Indeed,
all $\phi_i$'s do satisfy this \emph{stability} condition, thus
lending the validity to the proof.
